---
title: "Reproducible Research: Course Project 1"
author: "Steve Orosz"
date: "August 19, 2018"
output: html_document
---

```{r load data,  echo=TRUE}
knitr::opts_chunk$set(echo = TRUE)

#Libraries

library(lubridate)
library(dplyr)

#Set working directory

setwd("C:/Users/smoro/OneDrive/Desktop/Data Science/John Hopkins University/Reproducible Research/Project_One")

#getwd()

#Get file

Mydata <- read.csv("C:/Users/smoro/OneDrive/Desktop/Data Science/John Hopkins University/Reproducible Research/Project_One/activity.csv", header = TRUE)

head(Mydata)

#View(Mydata)

#Create weekday fields
MydataComplete <- Mydata %>% mutate (weekday = weekdays(as.Date(Mydata$date))) %>% mutate(weekdayNum = wday(Mydata$date))

#View(MydataComplete)

#Create data set without misssing values and with missing values

WithOut <-  na.exclude(MydataComplete)
With <- MydataComplete

#View(WithOut)
#View(With)


```


```{r Number of Steps per Day , echo=TRUE}

#Count total steps per day
ts <- tapply(WithOut$steps, WithOut$date , sum)

TotalStepDay <- data.frame(date = names(ts), sum = ts)

#View(TotalStepDay)


```

```{r Histogram of total steps per day , echo=TRUE}

#Histogram of total steps per day
hist(TotalStepDay$sum)

```

##What is mean and median total number of steps taken per day?
```{r Mean and median number of Steps per Day, echo=TRUE}

#Mean total steps per day
m1 <- tapply(WithOut$steps, WithOut$date , mean)

MeanStepDay <- data.frame(date = names(m1), DateMean = m1)

#View(MeanStepDay)

MeanStepDay

#View(MeanStepDay) Removed 0 steps to calculate median
preMedianStepDay <- WithOut %>% filter(WithOut$steps > 0)


#Median total steps per day
md <- tapply(preMedianStepDay$steps, preMedianStepDay$date , median)

MedianStepDay <- data.frame(date = names(md), DateMedian = md)

MedianStepDay

#View(MedianStepDay)

```
##What is the average daily activity pattern?
###Saturday has the highest step average than the rest of the week
```{r Average daily activity pattern, echo=TRUE}


FiveInterval <- WithOut  %>% mutate(FiveInterval = ((steps / interval) * 1.0) * 5)

m2 <- tapply(FiveInterval$steps, FiveInterval$weekdayNum,  mean)

MeanWeekDays <- data.frame(day = names(m2), mean = m2)

#Display time series of average steps taken over week

plot(  MeanWeekDays$mean, MeanWeekDays$day, type = "l", xlab = "Average Steps per Day", ylab = "Day of Week")


```
##Missing values Section
#What is the total number of rows with missing values?
```{r Missing Values, echo=TRUE}

sum(is.na(With$steps))

```
#Calculate mean and median per interval of steps. This will be used for days with NA for steps
#Date could not be used for NA data since no step data for them.

```{r Fill in missing Values with Mean, echo=TRUE}

#produce mean average
m3 <- tapply(WithOut$steps, WithOut$interval , mean)

MeanStepInterval <- data.frame(interval = names(m3), IntervalMean = m3)

#produce median average

m4 <- tapply(WithOut$steps, WithOut$interval , median) 

MedianStepInterval <- data.frame(interval = names(m4), IntervalMedian = m4)

MedianStepInterval

#Calculate Final Mean average for each day
CombinedFile <- merge(With, MeanStepInterval, by = "interval")
CombinedFile <- merge(CombinedFile, MeanStepDay, by = "date")
CombinedFile <- mutate(CombinedFile, FinalMean = ifelse(is.na(CombinedFile$DateMean), 
            CombinedFile$IntervalMean, CombinedFile$DateMean))

#Calculate Final Median average for each day
CombinedFile <- merge(CombinedFile, MedianStepInterval, by = "interval")
CombinedFile <- merge(CombinedFile, MedianStepDay, by = "date")
CombinedFile <- mutate(CombinedFile, FinalMedian = ifelse(is.na(CombinedFile$DateMedian), 
            CombinedFile$IntervalMedian, CombinedFile$DateMedian))

#View(CombinedFile)


```


```{r Number of Steps per Day for all data , echo=TRUE}

#Count total steps per day
ts2 <- tapply(With$steps, With$date , sum)

TotalStepDayAll <- data.frame(date = names(ts2), sum = ts2)

TotalStepDayAll<- mutate(TotalStepDayAll, FinalSum = ifelse(is.na(TotalStepDayAll$sum), 
            0, TotalStepDayAll$sum))



#View(TotalStepDayAll)



```
##The frequency of total steps per day with NA data that was imputed shows more frequency counts in 0 - 5000 step range than before. The rest of buckets are identical.

```{r Histogram of total steps per day for all data , echo=TRUE}

#Histogram of total steps per day
hist(TotalStepDayAll$FinalSum, main = "Histogram of total steps per day for all data")


hist(TotalStepDay$sum, main = "Histogram of total steps per day for only complete data")

```

```{r Compare weekdays to weekends , echo=TRUE}

#CombinedFile

#Create weekend subset 
SatSun <- c("Saturday", "Sunday")
weekend <- subset(CombinedFile, weekday %in% (SatSun))

#Create weekday subset
Wkdays <- c("Monday","Tuesday","Wednesday","Thursday","Friday")
weekdays <- subset(CombinedFile, weekday %in% (Wkdays))



```

```{r Five minute interval for all data, echo=TRUE}


#Get the 5 minute interval mean for NA rows
naSet <- CombinedFile[is.na(CombinedFile$steps),] %>% select(steps,date,interval, weekday, weekdayNum, FinalMean)
#naFiveInterval <- subset(naSet, interval == 5)

#Calculate five minute interal for NA file
naFiveInterval <- naSet  %>% mutate(FiveInterval = ((FinalMean / interval) * 1.0) * 5) %>% select(steps,date,interval, weekday, weekdayNum, FiveInterval)

#Combine the two five interval files into one
CombinedFiveInterval <- rbind(FiveInterval, naFiveInterval)

CombinedFiveIntervalFinal  <- na.exclude(CombinedFiveInterval) 
CombinedFiveIntervalFinal <- subset(CombinedFiveIntervalFinal, interval > 0 )


#Create weekday subset
Wkdays <- c("Monday","Tuesday","Wednesday","Thursday","Friday")
weekdays <- subset(CombinedFiveIntervalFinal, weekday %in% (Wkdays))

w1 <- tapply(weekdays$steps, weekdays$weekdayNum,  mean)

weekdayAverage <- data.frame(day = names(w1), mean = w1)

#View(weekdayAverage)

#Create weekend subset 
SatSun <- c("Saturday", "Sunday")
weekend <- subset(CombinedFiveIntervalFinal, weekday %in% (SatSun))

w2 <- tapply(weekend$steps, weekend$weekdayNum,  mean)

weekendAverage <- data.frame(day = names(w2), mean = w2)

#View(weekendAverage)




#Display time series of average steps for week days and then weekend days
par(mfrow=c(1,2))

plot(weekdayAverage$mean, weekdayAverage$day, type = "l", xlab = "Average Steps per Day", ylab = "Day of Week", main = "Weekday average steps all data")


plot(weekendAverage$mean, weekendAverage$day, type = "l", xlab = "Average Steps per Day", ylab = "Day of Week", main = "Weekend average steps all data")


```